Identifying covariate traits of trance and possession phenomena in the Ethnographic

### 2022-10-23 Readme.Rmd
fixed var coding, pls fix cor plot

# To do: one-hot encoding ended up creating unidentifiable vars (where both values of a col are present) so do something with that.

@@

cor(cord,use="pairwise.complete.obs")

@@

We compare this with how our best model categorises each society. This is our model's confusion matrix. `0` is absence of possession trance, `1` is presence of possession trance. The rows add up to values in the original, the columns are the model's predictions. The last two columns show the error rate. The model's accuracy on absence of possession trance is relatively low, though higher than chance (54 out of 80, p < 0.01 using a binomial test with a true p of .529). It is mostly accurate capturing the presence of possession trance.

@@

# precision = true 1/false 1
# recall = true 1/true1 + false0
precision = 68 / 94
recall = 68 / 71
f = (2 * precision * recall) / (precision + recall)
```

The model has an F-measure of `r round(f,2)`, with a penalty on overgeneralising possession trance. Though comparisons are hard to make, this is not particularly accurate for a binary categorisation problem. We did expect this, though, as it is unlikely that any complex cultural phenomenon could be reduced to a small number of cross-cultural predictors.

@@

# precision = true 1/false 1
# recall = true 1/true1 + false0
precision2 = 221 / 389
recall2 = 221 / 281
f2 = (2 * precision2 * recall2) / (precision2 + recall2)

@@

The figure below shows the proportion of societies with possession trance (dark blue) and without (light blue) across the ordered levels of the predictors that the model thinks are important. They come in order of how important they are, according to the model; prevalence of slavery, then family organisation, and so on. The predictor levels are ordered according to coding and generally go from less prevalent (top) to more prevalent (bottom):

For the most important variables, we see a clear correllation between the presence of possession trance and the variable scaling up. Societies in which slavery is more prevalent have more possession trance than those where it is less prevalent. Same is true for societies with more complex family structures, more social complexity, more complex settlement patterns, and larger settlements. The societal variables that do not, strictly speaking, measure social organisation are less obvious.

@@

These results go some way in confirming hypotheses on the cross-cultural correlates of possession trance in the literature. @bourguignon1968 explicitly tests the variables of slavery, jurisdictional hierarchy, settlement patterns, and kin groups. @shaara1992 point to a high value placed on female virginity at marriage, captured here by norms of premarital sexual behaviour for girls. These are all important variables in our model. @greenbaum1973 and @shaara1992 point to sexual division of labour, while @wood2018complexity to the role of moralising high gods, neither of which is relevant in our model.

@@

| Relationship|Reference|Our model|
|:-|:-|:-|
|less complexity$\rightarrow{}$trance w/o possession|@bourguignon1968|the same measures of complexity are relevant|
|women's social participation|@bourguignon1973religion|labour divisions are not relevant|
|increased role differentiation and slavery practices$\rightarrow$trance and possession|@greenbaum1973|slavery, social and class complexity are relevant|
|main mode of subsistence|@bourgouignon1977|not relevant|
|increased social participation$\rightarrow{}$possession|@swanson1978|indirectly, through domestic and social complexity|
|more complexity$\rightarrow{}$trance w/ possession|@winkelman1986|yes, see above|
|cognatic kinship$\rightarrow{}$more possession and trance|@shaara1992|not directly, but kinship structure is relevant|
|decreased social participation$\rightarrow{}$possession|@douglas2004natural|yes, see above|
|monotheism|@wood2018complexity|not relevant in our model|

>>>>

@@

# cor(cord,use="pairwise.complete.obs")

@@

We compare this with how our best model categorises each society. This is our model's confusion matrix. `0` is absence of possession trance, `1` is presence of possession trance. The rows add up to values in the original, the columns are the model's predictions. The last two columns show the error rate.

@@

# precision = true 1/tru1 + false 1
# recall = true 1/true1 + false0
precision = 64 / (64+12)
recall = 64 / (64+7)
f = (2 * precision * recall) / (precision + recall)
chisq.test(c(c(68,12),c(7,64)))
```

The model has an F-measure of `r round(f,2)`, with a penalty on slightly overgeneralising possession trance. Though comparisons are hard to make, this is not particularly accurate for a binary categorisation problem. We did expect this, though, as it is unlikely that any complex cultural phenomenon could be reduced to a small number of cross-cultural predictors.

@@

# precision = true 1/true1+false 1
# recall = true 1/true1 + false0
precision2 = 194 / (194+113)
recall2 = 194 / (194+87)
f2 = (2 * precision2 * recall2) / (precision2 + recall2)
chisq.test(c(c(264,113),c(87,194)))
```

@@

The figure below shows the proportion of societies with possession trance (dark blue) and without (light blue) across the ordered levels of the predictors that the model thinks are important. They come in order of how important they are, according to the model; prevalence of slavery, then family organisation, and so on. The predictor levels are ordered according to coding and generally go from a pattern being less prevalent (top) to more prevalent (bottom):

Ee see a clear correlation between the presence of possession trance and the variable scaling up, e.g. societies in which slavery is more prevalent have more possession trance than those where it is less prevalent. The across-level distributions are also interesting, e.g. the main distinction for patterns of slavery is for societies without slavery (first factor level) and all others with some form of slavery (subsequent factor levels).

@@

These results go some way in confirming hypotheses on the cross-cultural correlates of possession trance in the literature. @bourguignon1968 explicitly tests the variables of slavery, jurisdictional hierarchy, settlement patterns, and kin groups. @shaara1992 point to a high value placed on female virginity at marriage, captured here by norms of premarital sexual behaviour for girls, @wood2018complexity point to the role of moralising high gods. These are all important variables in our model. @greenbaum1973 and @shaara1992 point to sexual division of labour but this is not relevant in our model.

Note that @shaara1992 use cognatic kinship as a proxy of social complexity (which in itself is robustly supported by our model) while the measure referenced by @wood2018complexity, the presence of moralising high gods, has in itself been argued to correlate with social complexity (see @botero2014ecology and @watts2015broad).

@@

| Relationship|Reference|Our model|
|:-|:-|:-|
|less complexity$\rightarrow{}$trance w/o possession|@bourguignon1968|the same measures of complexity are relevant|
|women's social participation|@bourguignon1973religion|labour divisions are not relevant|
|increased role differentiation and slavery practices$\rightarrow$trance and possession|@greenbaum1973|slavery, social and class complexity are relevant|
|main mode of subsistence|@bourgouignon1977|not relevant|
|increased social participation$\rightarrow{}$possession|@swanson1978|indirectly, through domestic and social complexity|
|more complexity$\rightarrow{}$trance w/ possession|@winkelman1986|yes, see above|
|cognatic kinship$\rightarrow{}$more possession and trance|@shaara1992|not directly|
|decreased social participation$\rightarrow{}$possession|@douglas2004natural|yes, see above|
|monotheism|@wood2018complexity|yes|
